### 2018-05-27 write-up/abstract.tex
Finish Conclusion

\newpage
{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
\vspace{24pt}

In this thesis, I will argue that linear types are an appropriate,
\emph{type-based formalism} for expressing aliasing, read/write permissions,
memory allocation, re-use and deallocation, first, in the context of the APIs
of linear algebra libraries, consequently extending these principles to the
problem of \emph{efficient} matrix-expression compilation.

I will show that framing the problem using linear types can help \emph{reduce
bugs} by making precise and explicit the informal, ad-hoc practices typically
employed by human experts and linear algebra \emph{compilers} and automate
checking them.

As evidence for this argument, I will show programs written with this safety,
precision and explicitness (1) can be just as pleasant and convenient for a
programmer as less efficient, but higher-level linear algebra libraries and (2)
perform just as \emph{efficiently and predictably} as lower-level, less
readable and more error-prone interfaces linear algebra libraries. I will
further show that support for such programs can be implemented \emph{as
libraries} on top of existing programming languages and libraries.

\textbf{Summarise evaluation.}

\textbf{Summary future work.}

>>>>

\newpage
{\normalfont\Huge\sffamily\bfseries Abstract}
\vspace{24pt}

In this thesis, I argue that linear types are an appropriate, \emph{type-based
formalism} for expressing aliasing, read/write permissions, memory allocation,
re-use and deallocation, first, in the context of the APIs of linear-algebra
libraries and then in the context of matrix-expression compilation.

I show that framing the problem using linear types can help \emph{reduce bugs}
by making precise and explicit the informal, ad-hoc practices typically
employed by human experts and linear-algebra \emph{compilers} and automate
checking them.

As evidence for this argument, I show non-trivial, yet readable, linear-algebra
programs, that are safe and explicit (with respect to aliasing, read/write
permissions, memory allocation, re-use and deallocation) which (1) are more
memory-efficient than equivalent programs written using high-level
linear-algebra libraries and (2) perform just as predictably as equivalent
programs written using low-level linear-algebra libraries. I also argue
\emph{the experience} of writing such programs with linear types is
qualitatively better in key-respects. In addition to all of this, I show that
it is possible to provide such features \emph{as a library} on top of existing
programming languages and linear-algebra libraries.

### 2018-05-27 write-up/background.tex
Finish Conclusion

\prechapter{%
    I will outline the concept of linear types and show how they can be used to
    solve the problems faced by programmers writing code using linear-algebra
    libraries. I will emphasise \emph{practical} and intuitive explanations of
    linear types to keep this thesis accessible to working programmers as well
    as academics not familiar with type-theory and will give only a terse
    overview of the history and theory behind linear types for the interested
    reader to pursue further.
}%

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program-invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seems restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so too do typing rules make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once.\footnote{This definition may differ from more colloquial
uses in discussions surrounding \emph{substructural} type systems and/or
Rust.} While this restriction may seem limiting at first, precisely these
constraints can be used to express common invariants of the programs written by
working programmers every day. For example: a file or a socket, once opened
\emph{must} closed; all memory that is manually allocated \emph{must} be freed.
C++'s destructors and Rust's Drop-trait (and more generally, its
borrow-checker) attempt to enforce these constraints by basically doing the
same thing: any resource that has not been moved is deallocated at the end of
the current lexical scope. Notably, these languages also permit aliasing,
alongside rules enforcing when it is acceptable to do so.  On face value, the
above one-line description of linear types prevents aliasing or functions such
as $\lambda x.\;x \times x$, such features are still allowed (albeit in a more
restricted fashion) in a \emph{usable} linear type system designed for working
with linear-algebra libraries.

\section{Problem in Detail}
Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and contrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.} Most of what follows applies to \emph{dense}
linear-algebra computations rather than \emph{sparse} because memory
allocated for results typically depends on the sparsity of the inputs and
so is not immediately amenable to the techniques proposed in this thesis.

\subsection{One Too Many Copies and a Thousand Bytes Behind}
\begin{figure}[tbp]
    \centering
    \begin{minted}[fontsize=\small]{python}
# Numpy (Python)
import numpy.matlib
a = [[1,0],[0,1]]
b = [[4,1],[2,2]]
c = numpy.matmul(a,b)
    \end{minted}
    \begin{minted}[fontsize=\small]{julia}
# Julia
c = [1 0; 0 1] * [4 1; 2 2]
    \end{minted}
    \begin{minted}[fontsize=\small]{ocaml}

(* Owl (OCaml) *)
open Owl
let a = Mat.of_arrays [|[|1.;0.|];[|0.;1.|]|]
let b = Mat.of_array [|[|4.;1.|];[|1.;2.|]|]
let c = Mat.(a *@ b)
    \end{minted}
    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
\end{figure}

In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
\emph{value-semantic}, meaning that objects are \emph{values} just like
integers and floating-point numbers. This approach confers two key advantages to the
programmmer: it is easy to read (equational and algebraic declarations) and it
is easy to reason about (as one would with a mathematical formula). Although this
approach does permit \emph{aliasing}, the consequences are benign because the
result of any computation is a \emph{new} value, distinct from any used during
the calculation of that value.

@@

However, these advantages come with some costs: constantly producing new values
is wasteful on memory (although the example given in
Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
datasets can contain up to gigabytes of data). A complex expression may create
many short-lived temporaries which would need to be reclaimed by a
garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
\emph{value-semantic} approach offer a dichotomy for a user wishing to
implement a new algorithm: either use the existing and safe primitives to build
an easy to reason about but slower, more memory-intensive algorithm, or use
escape-hatches (typically provided by most libraries, which permit in-place
modification of objects) to build faster, and more efficient algorithms which
are harder to reason about.

\subsection{IHNIWTLM}
The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
illustrates the problem with the C++/FORTRAN side: legibility (and ease of
reasoning) is sacrificed at the altar of performance and efficiency.

Although escape-hatches do exist in value-semantic libraries, their use is
discouraged. Systematic consideration of performance requires lowering the
level of abstraction a programmer is working on. At this level, several
factors such as memory layout, allocation, re-use as well as cache behaviour
and parallelism become apparent. Of these, memory allocation and re-use are
of most relevance to linear-types and this thesis.

@@

\section{Proposed Solution}

My proposed solution to this dichotomy is a \emph{domain-specific language}
(DSL), called LT4LA (Linear Types for Linear Algebra). It offers readable,
explicit management of aliasing, read/write permissions, memory allocation,
re-use and deallocation all with automated checking: offering a safety net to
catch the baby whilst swiftly disposing of the bath water.  Although for
expository and testing purposes I have defined a concrete-syntax, a full
implementation would make use of a language's syntax-extension features (such
as PPX for OCaml) to \emph{embed} the DSL into the host language. Such an
embedding is straightforward but fairly tedious to implement. As a half-way
point, I used compile-time code generation to make the DSL's output available
to OCaml for testing and evaluation.

@@

\begin{figure}[tp]
    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
routine which takes two read-only matrices and a third matrix it can write to
and performs $C := AB + C$. How would we use such a routine to square a matrix?
Surely this would break linearity, since $A$ would be have to be given to the
function twice, and we can only do so once?

We can use a special primitive called \ltfla{share} to produce read-only
\emph{aliases} of any matrix (including ones we cannot write to). We then pass
this into the function, and it works as expected. Once the squared matrix has
been obtained, we may not want the original any more, and thus decide to free
it. Before we do so, we must first \ltfla{unshare} any read-only aliases that
exist -- in this context, \ltfla{unshare} swallows them (the only primitive
which can return fewer read-only values than it receives) and gives back a
single, read-write handle.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA.}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error show in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let () = free a1 in c
(* Error:                                                        *)
(* Could not show equality:                                      *)
(*     z arr                                                     *)
(* with                                                          *)
(*     z s arr                                                   *)
(*                                                               *)
(* Could not show z and z s are equal.                           *)
(* Are you trying to write to/free an array before unsharing it? *)
(* In test.lt, at line: 3 and column: 17                         *)
    \end{minted}
    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
\end{figure}

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}, emphasising small, but illustrative and
practical code examples, leaving \emph{what} is going on and \emph{why} it
works as details for the next chapter. I have shown that it is possible to have
high-level, explicit and automatically checked management of aliasing,
read/write permissions, memory allocation, re-use and deallocation with a
\emph{type system} -- preventing a whole class of errors that can occur with
lower-level languages \emph{at compile time}, using the of context linear
algebra libraries as as a specific example. In the next chapter, I will show
that these features can be expressed and implemented in a language-independent
manner to be provided and used \emph{as a library}.

>>>>

\prechapter{%
    I will outline the concept of linear types and show how they can be used to
    solve the problems faced by programmers writing code using linear-algebra
    libraries. I will emphasise \emph{practical} and intuitive explanations of
    linear types to keep this thesis accessible to programmers as well as
    academics not familiar with type-theory and will give only a terse overview
    of the history and theory behind linear types for the interested reader to
    pursue further.
}%

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seem restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so too do typing rules make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once\footnote{This definition may differ from more colloquial uses
in discussions surrounding \emph{substructural} type systems and/or Rust.}
(Figure~\ref{fig:example_rules} shows two example typing rules for a typical,
linearly typed language). While this restriction may seem limiting at first,
precisely these constraints can be used to express common invariants of the
programs written by programmers every day. For example: a file or a
socket, once opened \emph{must} closed; all memory that is manually allocated
\emph{must} be freed.  C++'s destructors and Rust's Drop-trait (and more
generally, its borrow-checker) attempt to enforce these constraints by
basically doing the same thing: any resource that has not been moved is
deallocated at the end of the current lexical scope. Notably, these languages
also permit aliasing, alongside rules enforcing when it is acceptable to do so.
On face value, the above one-line description of linear types prevents aliasing
or functions such as $\lambda x.\; (x, x)$; such features are still allowed
(albeit in a more restricted fashion) in a \emph{usable} linear type system
designed for working with linear-algebra libraries.

\begin{figure}
    \centering
    \begin{minipage}{0.3\textwidth}
        \centering
    \begin{prooftree}
        \AxiomC{}
        \RightLabel{\textsc{Var}}
        \UnaryInfC{$\Gamma, x : t \vdash x : t ; \Gamma$}
    \end{prooftree}
    \end{minipage}%
    \begin{minipage}{0.6\textwidth}
        \centering
    \begin{prooftree}
        \AxiomC{$\Gamma_1 \vdash e_1 : t_1 ; \Gamma_2$}
        \AxiomC{$\Gamma_2 \vdash e_2 : t_2 ; \Gamma_3$}
        \RightLabel{\textsc{Pair}}
        \BinaryInfC{$\Gamma_1 \vdash (e_1, e_2) : t_1 \otimes t_2 ; \Gamma_3$}
    \end{prooftree}
    \end{minipage}
    \caption{A typing rule has the general form $\Gamma_{\textrm{in}} \vdash e ; \Gamma_{\textrm{out}}$.
    Typing a variable \emph{removes} it from the environment. Typing a pair
    requires that each component of the pair uses \emph{different} resources
    from the environment.}\label{fig:example_rules}
\end{figure}

\section{Problem in Detail}

Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and contrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.} Most of what follows applies to \emph{dense}
linear-algebra computations rather than \emph{sparse} because memory
allocated for results typically depends on the sparsity of the inputs and
so is not immediately amenable to the techniques proposed in this thesis.

\subsection{One Too Many Copies and a Thousand Bytes Behind}
\begin{figure}[tbp]
    \centering
    \begin{minipage}{0.45\textwidth}
    \centering
    \begin{minted}[fontsize=\small]{python}
# Numpy (Python)
import numpy.matlib
a = [[1,0],[0,1]]
b = [[4,1],[2,2]]
c = numpy.matmul(a,b)
    \end{minted}
    \begin{minted}[fontsize=\small]{julia}
# Julia
c = [1 0; 0 1] * [4 1; 2 2]
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.45\textwidth}
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
(* Owl (OCaml) *)
open Owl
let a = Mat.of_arrays
  [|[|1.;0.|];[|0.;1.|]|]
let b = Mat.of_arrays
  [|[|4.;1.|];[|1.;2.|]|]
let c = Mat.(a *@ b)
    \end{minted}
    \end{minipage}
    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
\end{figure}

In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
\emph{value-semantic}, meaning that objects are \emph{values} just like
integers and floating-point numbers. This approach confers two key advantages to the
programmmer: it is easy to read (equational and algebraic expressions) and it
is easy to reason about (as one would with a mathematical formula). Although this
approach does permit \emph{aliasing}, the consequences are benign because the
result of any computation is a \emph{new} value, distinct from any used during
the calculation of that value.

@@

However, these advantages come with some costs: constantly producing new values
is wasteful on memory (although the example given in
Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
datasets can contain up to gigabytes of data). A complex expression may create
many short-lived temporaries which would need to be reclaimed by a
garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
\emph{value-semantic} approach offer a dichotomy for a user wishing to
implement a new algorithm: either use the existing and safe primitives to build
an easy to reason about but slower, more memory-intensive algorithm, or use
escape-hatches (typically provided by most libraries, which permit in-place
modification of objects) to build faster, and more memory-efficient algorithms
which are harder to reason about.

\subsection{IHNIWTLM}
The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
illustrates the problem with the C++/FORTRAN side: readability (equational
and algebraic expressions) and ease of reasoning is sacrificed at
the altar of performance and efficiency.

Although escape-hatches do exist in value-semantic libraries, their use is
discouraged. Systematic consideration of performance requires lowering the
level of abstraction a programmer is working on. At this level, several
factors such as memory layout, allocation, re-use as well as cache behaviour
and parallelism become apparent. Of these, \textbf{memory allocation and
re-use} are relevant to linear-types and this thesis.

@@

\section{Proposed Solution}\label{sec:prop_sol}

My proposed solution to this dichotomy (readability, ease of reasoning and
safety versus memory-efficiency) is a \emph{domain-specific language} (DSL),
called LT4LA (Linear Types for Linear Algebra). It is written in OCaml and
transpiles to OCaml. It offers readable, explicit management of aliasing,
read/write permissions, memory allocation, re-use and deallocation all with
automated checking: offering a safety net to catch the baby whilst swiftly
disposing of the bath water.  Although for expository and testing purposes I
have defined a concrete-syntax, a full implementation would make use of a
language's syntax-extension features (such as PPX for OCaml) to \emph{embed}
the DSL into the host language. Such an embedding is straightforward but fairly
tedious to implement. As a half-way point, I used compile-time code generation
to make the DSL's output available to OCaml for testing and evaluation.

@@

\begin{figure}[p]
    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
it can write to and performs $C := AB + C$. How would we use such a routine to
square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
since $A$ would be have to be given to the function twice, and we can only do
so once?

We can use a special primitive called \ltfla{share} to produce (more) read-only
\emph{aliases} of any matrix. We then pass these into the function, and it
works as expected. Once the squared matrix has been obtained, we may not want
the original any more, and thus decide to free it. Before we do so, we must
first \ltfla{unshare} any read-only aliases that exist -- in this context,
\ltfla{unshare} swallows them and gives back a single, read-write handle.

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
        0).}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error shown in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

Another way in which `simple\_dgemm' may be misused is by instantiating its
informal description of $C := AB + C$ with $B = C = A$ and so mistakenly
concluding that it computes $A := A^2 + A$ in-place. However, the type of
\ltfla{share} prevents this as well -- \ltfla{let (a11, a12) = share _ a1 in
simple_dgemm _ a11 _ a12 a2} would result in an error similar to the one shown
in Figure~\ref{fig:ltfla_free}.

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let () = free a1 in c
(* Error:                                                        *)
(* Could not show equality:                                      *)
(*     z arr                                                     *)
(* with                                                          *)
(*     z s arr                                                   *)
(*                                                               *)
(* Could not show z and z s are equal.                           *)
(* Are you trying to write to/free an array before unsharing it? *)
(* In test.lt, at line: 3 and column: 17                         *)
    \end{minted}
    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
\end{figure}

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}~\cite{boyland}, emphasising small, but
illustrative and practical code examples, leaving \emph{what} is going on and
\emph{why} it works as details for the next chapter. I have shown that it is
possible to solve the dichotomy of readability, ease of reasoning and safety
versus memory-efficiency by providing explicit and automatically checked
management of aliasing, read/write permissions, memory allocation, re-use and
deallocation with a \emph{type system}. This prevents a whole class of errors
that can occur with lower-level languages \emph{at compile time}. I
demonstrated these features using the context of linear-algebra libraries as as
a specific example. In the next chapter, I will show that these features can be
further applied to the problem of \emph{matrix expression compilation}. I will
also explain how to express and implement them so that they can be provided and
used \emph{as a library}.

### 2018-05-27 write-up/conclusion.tex
Finish Background, continue Implementation
Start Evaluation
Finish Evaluation
Finish Conclusion

\chapter{TO DO: Conclusion}

\begin{guidance}
    As you might imagine: summarizes the dissertation, and draws
    any conclusions. Depending on the length of your work, and
    how well you write, you may not need a summary here.

    You will generally want to draw some conclusions, and point
    to potential future work.
\end{guidance}

\textbf{What are the benefits of the type-based approach I have taken? Have I successfully 
argued whatever I wrote in the introduction?}

@@

\textbf{Graph stuff... check with supervisors.}

All of the matrix expression compilers mentioned in previous section construct
some sort of dataflow graph to represent the computation being executed. While
this seems intuitive, there is no formal argument for this  appr
Some directions in which a type-based approach to efficient matrix expression compilation could
be taken are:

>>>>

\chapter{Conclusion}

% \begin{guidance}
%     As you might imagine: summarizes the dissertation, and draws
%     any conclusions. Depending on the length of your work, and
%     how well you write, you may not need a summary here.
%
%     You will generally want to draw some conclusions, and point
%     to potential future work.
% \end{guidance}

In this thesis, I presented \textbf{linear types with fractional-capabilities}
as a type-based formalism for expressing \textbf{aliasing, read/write
permissions, memory allocation, re-use and deallocation}; I provided a detailed
description of each feature of the type-system, its intent and implementation
in Section~\ref{sec:core_lang}.

I used that formalism to make \textbf{precise and explicit the intensional
behaviour and assumptions} (about aliasing, read/write permissions and memory
re-use) of linear-algebra library APIs, first with the simplified example of
the fictional `simple\_dgemm' primitive in Section~\ref{sec:prop_sol} -- where
I showed how \textbf{the types made it impossible to use `simple\_dgemm'
incorrectly} -- and then, more realistically with the types of LT4LA primitives
(including actual CBLAS/LAPACKE routines) listed in
Appendix~\ref{chap:primitives}.

I also used that formalism to implement a simple, yet effective matrix
expression `compiler' based on pattern-matching. I was able to use to that
compiler to write, a \textbf{non-trivial, yet readable} (equational and
algebraic) linear-algebra program (Kalman filter, as described in
Section~\ref{sec:matrix_exps}). I shared how, in doing so, I could use the
type-checking as a tool to \textbf{automatically highlight bugs} regarding
aliasing, read/write permissions and memory re-use of a reference
implementation (Section~\ref{sec:qual_benefits}) and the matrix expression
compiler that generated it (Section~\ref{sec:mat_exp_comp}).

% This suggests that although it is tempting to think of matrix expression
% compilation as a data-flow problem in terms of graphs, such an approach may
% struggle to accurately capture the intensional behaviour and assumptions of
% linear-algebra kernels which seem to be expressed nicely with linear types and
% fractional-capabilities.

I showed that this implementation was \textbf{more memory-efficient} than one
written using a high-level linear-algebra libraries and performed just as
\textbf{predictably} as one written using a low-level linear-algebra library
(Chapter~\ref{chap:eval}). The type system ensured that the implementation was
safe and explicit with respect to aliasing, read/write permissions, memory
allocation, re-use and deallocation.

Lastly, I was able to provide all of this functionality as a \textbf{usable
OCaml library}, by generating \textbf{readable and not obviously safe} OCaml
code (1) which looks like it is written by a human/expert and (2) whose types
reflect its intensional behaviour and assumptions (interpreting types into
OCaml, Figure~\ref{fig:type_grammar}). This library, LT4LA, produces
\textbf{helpful type-errors} (Section~\ref{sec:prop_sol}), comes with an
interactive REPL and is documented and well-tested
(Section~\ref{sec:structure_lt4la}).

@@

LT4LA is a proof-of-concept design for linearly-typed DSL for writing
linear-algebra programs in a way that can work with existing languages and
libraries. It is made of up of the core language and its matrix expression
`compiler'; both could be extended in many ways.

\subsection{Core Language}

Although I use well-established type system features, it may be worthwhile to
formally state and verify my claimed safety properties.  The logical conclusion
of LT4LA's core language would be a language expressive enough to provide a
usable, linearly-typed interface to the unsafe parts of the Owl library for
OCaml, perhaps as a functor like its current, `Lazy' interface.  To do that,
the language implementation would need to support more than just 64-bit
floating-point numbers for elements (for example, 16-bit, 32-bit or complex),
polymorphism, a unified way of dealing with arrays and matrices (perhaps
through sub-typing). It may also be useful to explore localised linearity
\emph{inference}, to automate the reasoning I did manually when implementing a
memory-efficient Kalman filter.

Another, equally valid but opposing direction may be to decrease the amount of
expressivity, to bring it closer to Fortran and C so that it would be easier to
emit code in those languages. A full set of linearly-typed BLAS/LAPACK bindings
(to encode information in documentation at the type level) would be a good
first step, perhaps leading into more research on formal semantics for Fortran.

\subsection{Matrix Expression Compilation}

All of the matrix expression compilers mentioned in the previous chapter
construct some sort of data-flow graph to represent the computation being
executed. While this seems intuitive, there is no formal argument for this
approach.  Some directions in which a type-based approach to efficient matrix
expression compilation could be taken are:
